The Gillespie Algorithm: Stochastic Modeling of Molecular Reaction Networks

### 2024-12-02 paper/Gillespie-Algorithm.Rmd
final push!
grammar edits

---
title: "The Gillespie Algorithm: Stochastic Modeling of Molecular Reaction Networks"
author: 
  - Jiayu Guo
  - Sinan Ma 
  - Anna Szydlowski
thanks: "We would like to thank Dr. Omidali A. Jazifor for his invaluable guidance throughout this project. The source code for simulations can be found on our [GitHub repository](https://github.com/masinan123/Gillespie-algorithm.git)."
date: "`r Sys.Date()`"
abstract: "The Gillespie Algorithm provides a framework for simulating stochastic molecular reaction networks, capturing the inherent randomness of chemical and biochemical processes. Unlike deterministic methods, the algorithm accurately models systems with low molecular counts, where fluctuations significantly influence reaction dynamics. This project explores the theoretical foundation of the Gillespie Algorithm, including its basis in the master equation, and demonstrates its application in simulating two molecular systems: the photodimerization of thymine, a process linked to UV-induced DNA damage, and the radioactive decay of isotopes. Through these simulations, the algorithm's capability to model stochastic trajectories is highlighted, offering insights into reaction mechanisms and the role of randomness in molecular processes. The implications of these findings extend to a variety of disciplines, including molecular biology, environmental monitoring, and nuclear medicine."
output:
  pdf_document:
    number_sections: true
    citation_package: biblatex
linkcolor: blue
urlcolor: blue
spacing: double
header-includes:
  - \usepackage{fancyhdr}
  - \pagestyle{fancy}
  - \fancyhead[L]{The Gillespie Algorithm}
  - \fancyfoot[CO,CE]{}
  - \fancyfoot[LE,RO]{\thepage}
  - \usepackage{setspace}
  - \doublespacing

@@

The Gillespie Algorithm was presented in 1976 by Daniel T. Gillespie (@Gillespie1976). This algorithm was developed  to answer the question, “If a fixed volume V contains a spatially uniform mixture of N chemical species which can inter-react through M specified chemical reaction channels, then given the numbers of molecules of each species present at some initial time, what will these molecular population levels be at any later time?” (@Gillespie1977). Given the trajectory of this molecular reaction is a stochastic process, this algorithm can simulate various molecular equation systems, such as radioactive decay models or decaying-dimerization reaction sets (@GillespieSSA).

@@

In the following sections, we will first provide an overview of the Gillespie Algorithm and its historical development in the [Section 2](#sec-Topic-Review). Next, the theoretical underpinnings of the algorithm, including its basis in the master equation and stochastic modeling principles, will be detailed in the [Section 3](#sec-Theory). We will then demonstrate the application of the Gillespie Algorithm to two molecular systems in the [Section 4](#sec-application): the photodimerization of thymine, and the stochastic decay of radioactive isotopes. These simulations will be performed using the R programming language (@citeR). Finally, our [Section 5](#sec-conclusion) will summarize the findings, discuss the algorithm's broader implications, and explore potential future applications.

@@

The idea of master equation is inherently stochastic. Firstly $P(S,t)$ itself is a probabilistic description of the system. Next, in most of the cases, the reaction/process is stochastic. Therefore the master equation describes the ensemble average of all possible systems, here `ensemble` is the collection of all possible states with the same initial condition and macroscopic parameters. Following some certain initial distribution, we are not guaranteed to reach the same distribution at time $t$.

In the article *Cellular growth and division in the Gillespie algorithm* (@Lu2004), the authors discussed the system of well-stirred mixture of $N$ chemical species. Since the system is well-stirred, the distribution of particles is always uniform in space. The state of the system can be described using a N-vector $Y=(X_1,...,X_N)$, where $X_i$ is the number of molecules of type $i$ in the system. Assume there are $\mu=1,...,M$ elementary reaction channels $R_{\mu}$. Let $c_{\mu}$ be the reaction rate, i.e., the probability that a random combination of molecules from channel $R_{\mu}$ selected at the moment $t$ react in the interval $[t,t+dt)$ with probability $c_{\mu}dt+o(dt)$. Let $h_{\mu}(Y)$ be the total number of possible distinct combinations of molecules for a channel $R_{\mu}$ when the system is in state $Y$, and $\alpha_{\mu}=(\alpha_{1,\mu},...,\alpha_{N,\mu})$ is a constant stoichiometric vector prescribing the change in the state of the system after the reaction $R_{\mu}$ has occurred.

@@

With the above setting-up, The master equation for this system becomes $$\frac{\partial}{\partial t}P(Y,t|Y_0,t_0)=\sum_{\mu=1}^M c_{\mu}h_{\mu}(Y-\alpha_{\mu})P(Y-\alpha_{\mu},t|Y_0,t_0)-\sum_{\mu=1}^M c_{\mu}h_{\mu}(Y)P(Y,t|Y_0,t_0)$$ here $P(Y,t|Y_0,t_0)$ is the probability of the system being in state $Y$ at time $t$ given it is in state $Y_0$ at time $t_0$.

@@

The master equation is a coupled linear ordinary equation for $|S_Y|$ dependent variables $P(Y,t|Y_0,t_0)$, where $|S_Y|$ is the size of the state space. For a system with $n$ particles, $|S_Y|$ is of order $O(n^N)$. Therefore it is almost impossible to solve the equation analytically, especially when the number of possible elementary reactions $M$ is large.

@@

Assuming that each reaction $R_{\mu}$ is independent with rate $a_{\mu}=h_{\mu}c_{\mu}$, the first occurrence $\tau_{\mu}$ of the reaction $R_{\mu}$ follows the exponential distribution with rate/propensity $a_{\mu}$. The first occurrence time of some reaction is then $\tau=\min_{\mu}\tau_{\mu}$, it follows  the exponential distribution with an overall rate/propensity $A=\sum_{\mu}a_{\mu}$. Given that a reaction occurs, the probability that it is the $\mu'$-th reaction is $P(\mu=\mu')=\frac{a_{\mu'}}{A}$. Therefore the the original Gillespie recipe follows the following steps:

@@

Therefore, this simulation involves four species: pyrimidine monomers Thymine 1 (\( T_1 \)), Thymine 2 (\( T_2 \)), and dimer photoproducts \( CPD \) and \( 64PP \); and two reactions channels: \( c_1 \) and \( c_2 \).
\[
T_1 + T_2 \xrightarrow{c_1} CPD
\]
\[
T_1 + T_2 \xrightarrow{c_2} 64PP
\]

We will simulate the dimerization of thymine using propensity based on the formation yield of CPD (~37 per \(10^6\) normal bases) and 64-PP (~2 per \(10^6\) normal bases) found by exposing skin cells to a UVB dose of 0.2 \(J/cm^2\) (@Mouret2006). As the amount of thymine bases in DNA varies from person to person, we will simulate thymine photodimerization of 1000 pairs of thymine.

@@

As shown in Figure \ref{fig:dimerization}, both CPD and 64-PP formation follow logarithmic growth, which aligns with expectations based on specific UV radiation conditions and previously reported quantum yields (@MartinezFernandez2022; @Mouret2006). The number of reactions per unit time levels off as pyrimidine pairs decrease, which is consistent with the propensity functions, as the propensity of 64-PP formation is much significantly than the propensity of CPD formation (@Mouret2006). This plateau suggests that, as the number of available reactants decreases, the rate of reaction slows down.
Due to the stochastic nature of molecular reactions, this simulation demonstrates how the Gillespie algorithm reflects fluctuations of photodimerization as it models the probabilistic process of each individual reaction event. These fluctuations are especially important in low-molecule systems, where randomness significantly impacts the overall reaction dynamics, providing a more accurate representation of the molecular behavior compared to deterministic models.

@@

#### Implications and Real-World Applications
The Gillespie algorithm offers profound implications for understanding molecular reactions, particularly in systems where stochastic effects are prominent, such as the dimerization. This process, highlighted in Figure \ref{fig:dimerization}, is widely studied, as it can lead to mutations potentially resulting in skin cancer, when not repaired. The Gillespie  algorithm allows us to realistically model these reactions, as it demonstrates the probabilistic nature of DNA damage, capturing fluctuations in molecular interactions that deterministic models cannot, providing a deeper understanding of molecular events, insights into mutation rates, and the impact of environmental factors like UV radiation on genetic integrity. The algorithm's ability to model these stochastic processes is crucial in areas such as genetics, where random molecular events can significantly influence disease outcomes. 
This algorithm can be further adjusted to show molecular reactions with various molecular species and reaction channels. Using the Gillespie algorithm to model molecular reactions has wide-ranging implications in fields such as genetics, cancer research, pharmacology, environmental science, and biotechnology. By capturing the inherent randomness in molecular processes, this approach can provide valuable insights into complex biological systems, enabling better prediction of outcomes, such as in disease development, medicine, and environmental impact studies.

@@

The stochastic approach modeled here provides insights into systems with low molecule counts, where randomness plays a significant role. For example, in radiological dosimetry, small-scale decay dynamics can impact the precision of dose delivery. In environmental monitoring, understanding the stochastic nature of decay processes aids in the interpretation of measurement data, particularly in situations where low concentrations of isotopes are involved.

@@

The Gillespie Algorithm represents a significant role in the numerical simulation of stochastic molecular reaction networks. Its ability to generate statistically trajectories of reaction systems provides insights into processes governed by inherent randomness, especially in systems with low molecule counts. This project has illustrated the versatility and power of the algorithm through two applications: the photodimerization of thymine and the radioactive decay of isotopes.

@@

The implications of this work extend beyond the specific applications discussed. By providing a framework for understanding and simulating stochastic reaction systems, the Gillespie Algorithm continues to play a pivotal role in advancing our knowledge of molecular biology, chemical kinetics, and physical chemistry. Future developments in computational efficiency and hybrid methods may further expand its applicability, enabling the simulation of even more complex systems with greater precision.

@@

### Visulization for Figure \ref{fig:decay-linear}

@@

### Visulization for Figure \ref{fig:decay-log}

>>>>

---
title: "The Gillespie Algorithm: Stochastic Modeling of Molecular Reaction Networks"
author: 
  - Jiayu Guo
  - Sinan Ma 
  - Anna Szydlowski
thanks: "We would like to thank Dr. Omidali A. Jazi for his invaluable guidance throughout this project. The source code for simulations can be found on our [GitHub repository](https://github.com/masinan123/Gillespie-algorithm.git)."
date: "`r Sys.Date()`"
abstract: "The Gillespie Algorithm provides a framework for simulating stochastic molecular reaction networks by capturing the inherent randomness of chemical and biochemical processes. Unlike deterministic methods, the algorithm accurately models systems with low molecular counts where fluctuations significantly influence reaction dynamics. This project explores the theoretical foundation of the Gillespie Algorithm, including its basis in the master equation, and demonstrates its application in simulating two molecular systems: the photodimerization of thymine, a process linked to UV-induced DNA damage, and the radioactive decay of isotopes. These simulations highlight the algorithm’s capability to model stochastic trajectories and offer insights into reaction mechanisms and the role of randomness in molecular processes. The implications of these findings extend to a variety of disciplines, including molecular biology, environmental monitoring, and nuclear medicine."
output:
  pdf_document:
    number_sections: true
    citation_package: biblatex
linkcolor: blue
urlcolor: blue
spacing: double
header-includes:
  - \usepackage{fancyhdr}
  - \pagestyle{fancy}
  - \fancyhead[L]{The Gillespie Algorithm}
  - \fancyfoot[CO,CE]{}
  - \fancyfoot[LE,RO]{\thepage}
  - \usepackage{setspace}
  - \doublespacing

@@

The Gillespie Algorithm was presented in 1976 by Daniel T. Gillespie (@Gillespie1976). This algorithm was developed  to answer the question, “If a fixed volume V contains a spatially uniform mixture of N chemical species which can inter-react through M specified chemical reaction channels, then given the numbers of molecules of each species present at some initial time, what will these molecular population levels be at any later time?” (@Gillespie1977). Given that the trajectory of this molecular reaction is a stochastic process, this algorithm can simulate various molecular equation systems, such as radioactive decay models or decaying-dimerization reaction sets (@GillespieSSA).

@@

The following [Section 2](#sec-Topic-Review) first provides an overview of the Gillespie Algorithm and its historical development. Next, the theoretical underpinnings of the algorithm, including its basis in the master equation and stochastic modeling principles, are detailed in [Section 3](#sec-Theory). In [Section 4](#sec-application), we demonstrate the application of the Gillespie Algorithm to two molecular systems: the photodimerization of thymine, and the stochastic decay of radioactive isotopes. These simulations are performed using the R programming language (@citeR). Finally, [Section 5](#sec-conclusion) summarizes our findings, discusses the algorithm's broader implications, and explores potential future applications.

@@

The idea of master equation is inherently stochastic. Firstly $P(S,t)$ itself is a probabilistic description of the system. Next, in most cases, the reaction/process is stochastic. Therefore the master equation describes the ensemble average of all possible systems. In this context, `ensemble` is the collection of all possible states with the same initial condition and macroscopic parameters. Following some certain initial distribution, we are not guaranteed to reach the same distribution at time $t$.

In the article *Cellular growth and division in the Gillespie algorithm* (@Lu2004), the authors discuss the system of well-stirred mixture of $N$ chemical species. Since the system is well-stirred, the distribution of particles is always uniform in space. The state of the system can be described using a N-vector $Y=(X_1,...,X_N)$, where $X_i$ is the number of molecules of type $i$ in the system. Assume there are $\mu=1,...,M$ elementary reaction channels $R_{\mu}$. Let $c_{\mu}$ be the reaction rate, i.e., the probability that a random combination of molecules from channel $R_{\mu}$ selected at the moment $t$ react in the interval $[t,t+dt)$ with probability $c_{\mu}dt+o(dt)$. Let $h_{\mu}(Y)$ be the total number of possible distinct combinations of molecules for a channel $R_{\mu}$ when the system is in state $Y$, and $\alpha_{\mu}=(\alpha_{1,\mu},...,\alpha_{N,\mu})$ is a constant stoichiometric vector prescribing the change in the state of the system after the reaction $R_{\mu}$ has occurred.

@@

With the above set-up, the master equation for this system becomes $$\frac{\partial}{\partial t}P(Y,t|Y_0,t_0)=\sum_{\mu=1}^M c_{\mu}h_{\mu}(Y-\alpha_{\mu})P(Y-\alpha_{\mu},t|Y_0,t_0)-\sum_{\mu=1}^M c_{\mu}h_{\mu}(Y)P(Y,t|Y_0,t_0)$$ here $P(Y,t|Y_0,t_0)$ is the probability of the system being in state $Y$ at time $t$ given it is in state $Y_0$ at time $t_0$.

@@

The master equation is a coupled linear ordinary equation for $|S_Y|$ dependent variables $P(Y,t|Y_0,t_0)$, where $|S_Y|$ is the size of the state space. For a system with $n$ particles, $|S_Y|$ is of order $O(n^N)$. It is therefore almost impossible to solve the equation analytically, especially when the number of possible elementary reactions $M$ is large.

@@

Assuming that each reaction $R_{\mu}$ is independent with rate $a_{\mu}=h_{\mu}c_{\mu}$, the first occurrence $\tau_{\mu}$ of the reaction $R_{\mu}$ follows the exponential distribution with rate/propensity $a_{\mu}$. The first occurrence time of some reaction is then $\tau=\min_{\mu}\tau_{\mu}$, it follows  the exponential distribution with an overall rate/propensity $A=\sum_{\mu}a_{\mu}$. Given that a reaction occurs, the probability that it is the $\mu'$-th reaction is $P(\mu=\mu')=\frac{a_{\mu'}}{A}$. Therefore, the the original Gillespie recipe follows the following steps:

@@

Therefore, this simulation involves four species: pyrimidine monomers Thymine 1 (\( T_1 \)) and Thymine 2 (\( T_2 \)), and dimer photoproducts \( CPD \) and \( 64PP \); and two reaction channels: \( c_1 \) and \( c_2 \).
\[
T_1 + T_2 \xrightarrow{c_1} CPD
\]
\[
T_1 + T_2 \xrightarrow{c_2} 64PP
\]

We simulate the dimerization of thymine using propensity based on the formation yield of CPD (~37 per \(10^6\) normal bases) and 64-PP (~2 per \(10^6\) normal bases) found by exposing skin cells to a UVB dose of 0.2 \(J/cm^2\) (@Mouret2006). As the amount of thymine bases in DNA varies from person to person, we will simulate thymine photodimerization of 1000 pairs of thymine.

@@

As shown in Figure \ref{fig:dimerization}, both CPD and 64-PP formation follow logarithmic growth, which aligns with expectations based on specific UV radiation conditions and previously reported quantum yields (@MartinezFernandez2022; @Mouret2006). The number of reactions per unit time levels off as pyrimidine pairs decrease, which is consistent with the propensity functions, as the propensity of 64-PP formation is markedly higher than the propensity of CPD formation (@Mouret2006). This plateau suggests that as the number of available reactants decreases, the rate of reaction slows down.
Due to the stochastic nature of molecular reactions, this simulation demonstrates how the Gillespie algorithm reflects fluctuations of photodimerization as it models the probabilistic process of each individual reaction event. These fluctuations are especially important in low-molecule systems, where randomness significantly impacts the overall reaction dynamics, providing a more accurate representation of the molecular behavior compared to deterministic models.

@@

#### Implications and Real-World Applications
The Gillespie algorithm offers profound implications for understanding molecular reactions, particularly in systems where stochastic effects are prominent, such as the dimerization. This process, highlighted in Figure \ref{fig:dimerization}, is widely studied as it can lead to mutations potentially resulting in skin cancer when not repaired. The Gillespie algorithm allows us to realistically model these reactions, as it demonstrates the probabilistic nature of DNA damage by capturing fluctuations in molecular interactions that deterministic models cannot, and provides a deeper understanding of molecular events, insights into mutation rates, and the impact of environmental factors like UV radiation on genetic integrity. The algorithm's ability to model these stochastic processes is crucial in areas such as genetics, where random molecular events can significantly influence disease outcomes. 
This algorithm can be further adjusted to show molecular reactions with various molecular species and reaction channels. Using the Gillespie algorithm to model molecular reactions has wide-ranging implications in fields such as genetics, cancer research, pharmacology, environmental science, and biotechnology. By capturing the inherent randomness in molecular processes, this approach can provide valuable insights into complex biological systems as it enables a better prediction of outcomes, such as in disease development, medicine, and environmental impact studies.

@@

The stochastic approach modeled in this project provides insights into systems with low molecule counts, where randomness plays a significant role. For example, in radiological dosimetry, small-scale decay dynamics can impact the precision of dose delivery. In environmental monitoring, understanding the stochastic nature of decay processes aids in the interpretation of measurement data, particularly in situations where low concentrations of isotopes are involved.

@@

The Gillespie Algorithm represents a significant role in the numerical simulation of stochastic molecular reaction networks. Its ability to generate statistical trajectories of reaction systems provides insights into processes governed by inherent randomness, especially in systems with low molecule counts. This project has illustrated the versatility and power of the algorithm through two applications: the photodimerization of thymine and the radioactive decay of isotopes.

@@

The implications of this work extend beyond the specific applications discussed. By providing a framework for understanding and simulating stochastic reaction systems, the Gillespie Algorithm continues to play a pivotal role in advancing our knowledge of molecular biology, chemical kinetics, and physical chemistry. Future developments in computational efficiency and hybrid methods may further expand its applicability and enable the simulation of even more complex systems with greater precision.

@@

### Visulization for Figure \ref{fig:decay-linear} and Figure \ref{fig:decay-log}

@@
